The demographic costs of sexually antagonistic selection in partially selfing populations

### 2022-01-03 doc/SA-Hermaphrodites-wDemography-R1.tex
Minor edits
Minor tweaks

%To persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1937}, \textcolor{blue}{rapid, unpredictable, or extreme changes} in environmental conditions \citep{Maynard-Smith1976, LandeShannon1996, OrrUnckless2008,Lyberger-etal-2021}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when \textcolor{blue}{different fitness components are negatively correlated} \citep{CharlesworthHughes2000, ConnallonHall2018}. 
\textcolor{blue}{A population's ability to adapt to its environment and persist in the long term depends upon the genetic variation that it carries \citep{Fisher1930, Lewontin1975}. Yet, populations are heterogenous collections of individuals existing in a complex world where adaptation can be impeded by conflicting selection and gene-flow between different classes of individuals, or more generally, when different fitness components are negatively correlated \citep{Prout2000, CharlesworthHughes2000, ConnallonHall2018}.}

@@

\textcolor{blue}{While significant efforts have been made to predict when SA polymorphisms should be maintained and to detect SA genetic variation from empirical data, less attention has been paid to the demographic consequences of SA selection \citep{MatthewsConnallon2019}}. Most population genetic theory, including SA selection theory, models the relative fitness of genotypes in populations of constant (and often infinite) size. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals \textcolor{blue}{which link genotype distributions to population dynamics (e.g., \citealt {johnston2013life, merot2020balancing}). For over 50 years, the field of eco-evolutionary dynamics has sought to jointly model the genetic and demographic processes involved in evolutionary change. Theoretical models combining population genetics with density-dependent population growth \citep{roughgarden1971density} and stage-structured populations (consolidated in \citealt{Charlesworth_1994}) paved the way toward understanding the demographic consequences of allele frequency changes due to selection. Various extensions of these theories have been made to study life-history evolution \citep{orive1995senescence, orive2001somatic}, spatial structure \citep{ronce2000kin}, kin selection \citep{rousset2004genetic}, and the evolution of quantitative traits \citep[e.g.,][]{coulson2006putting, barfield2011evolution, childs2016evolution, orive2017effects}. This growing body of eco-evolutionary theory makes plain that demographic processes must be taken into account explicitly to fully understand how genetic variation for fitness is shaped and maintained in natural populations.}

@@

\textcolor{blue}{Individuals contribute  gametes to the population outcrossing gamete pool according to their fertility, which depends on their stage and genotype at the SA locus. The allele frequencies in the male gamete pool are then calculated as 
\begin{linenomath*}
\begin{equation} \label{eq:maleGametePool}
	\left(
		\begin{array}{c}
			q^{\prime}_{A} \\
			q^{\prime}_{a} \\
		\end{array} \right) \propto 
				\left(
		\begin{array}{c}
			\mbf{1}^{\intercal}_{\omega}\mbf{F}^{\prime}_{AA}\mbf{p}^{\prime}_{AA}+ \frac{1}{2} \mbf{1}^{\intercal}_{\omega}  \mbf{F}^{\prime}_{Aa} \mbf{p}^{\prime}_{Aa}  \\
		\frac{1}{2} \mbf{1}^{\intercal}_{\omega}  \mbf{F}^{\prime}_{Aa} \mbf{p}^{\prime}_{Aa} + \mbf{1}^{\intercal}_{\omega}\mbf{F}^{\prime}_{aa}\mbf{p}^{\prime}_{aa}\\
		\end{array} \right),
\end{equation}
\end{linenomath*}
where equality is obtained by normalizing the frequencies such that they add to one (see Online Supplementary Materials for a derivation of Eq(\ref{eq:maleGametePool})). The first entry sums the  stage-specific contributions to the gamete pool of heterozygotes and $AA$ homozygotes; the second entry sums the stage-specific contributions to the gamete pool of heterozygotes and $aa$ homozygotes. }

\textcolor{blue}{ We now have all the ingredients to calculate the allele frequencies in the male gamete pool for the two-stage model in Eq(\ref{eq:F_male}), Eq(\ref{eq:maleGametePool}), and Table \ref{tab:Fitness}. Denoting the proportion of reproducing adults of genotype $i$ with $p^R_i$, the allele frequencies in the male gamete pool are given by 
\begin{linenomath*}
\begin{equation} \label{eq:maleGametePool_example}
\left(
		\begin{array}{c}
			q^{\prime}_{A} \\
			q^{\prime}_{a} \\
		\end{array} \right) = 
				\frac{1}{(1-s_m)p^R_{aa}+ (1-hs_m)p^R_{Aa}+p_{aa}}\left(
		\begin{array}{c}
(1-s_m)p^R_{AA}	+\frac{1}{2}(1-hs_m)p^R_{Aa}\\
p^R_{aa}+\frac{1}{2}(1-hs_m)p^R_{Aa}
		\end{array} \right).
\end{equation}
\end{linenomath*}
Note that this equation is identical to the classical population genetic result for allele frequencies after selection among males in a two-sex SA model, with the caveat that only reproductively active adults contribute to the gamete pool \citep[e.g., see Eq(1a) of][]{Kidwell1977}. Eq(\ref{eq:maleGametePool_example}) increasingly deviates from the classical population genetics result if additional reproductive adult age or stage classes are included (see Eq(\ref{eq:maleGametePool})).  }

@@

We identify conditions where SA polymorphism is 'protected' by evaluating the stability of populations initially fixed for either SA allele to invasion by the other (i.e., we assessed stability at the boundary equilibrium genotype frequencies of $\hat{\mbf{p}}_{AA} = 1$ and $\hat{\mbf{p}}_{aa} = 1$; \citealt{Levene1953, Prout1968, deVriesCaswell2019b}). The formal conditions for a protected polymorphism are determined by linearizing the model in the vicinity of the boundary equilibria ($\hat{\mbf{p}}$), and evaluating the magnitude of the largest eigenvalue of the Jacobian matrix of the linearization. A full derivation of the Jacobian and details of the invasion analysis are provided in the Online Supplementary Material, and the relevant leading eigenvalues are presented in \ref{App:Eigen}.

@@

Mortality from inbreeding depression quickly outweighs the beneficial effect of selfing on the equilbirium female load that was apparent in the previous section. In contrast to our earlier results, predominantly outcrossing populations are predicted to have the highest proportion of demographically viable polymorphic space when inbreeding depression is taken into acount (compare fig.~\ref{fig:polySpace} with fig.~\ref{fig:deltaPolySpace}). Regardless of the life-history stage at which inbreeding depression affects survival, populations with intermediate to high selfing rates are unlikely to harbour SA polymorphism unless they can afford the resulting loss of self-fertilized ovules/offspring. An interesting alternative interpetation of these results is that populations with intermediate to high selfing rates and population intrinsic growth rates near one are vulnerable to extinction if a sexually antagonistic allele invades the population.

@@

The first and central finding of our study is that when population intrinsic growth rates approach one, the deleterious effects of segregating male-beneficial SA alleles on female fecundity can result in extinction over much of the parameter space where SA polymorphism is maintained. Since intrinsic growth rates far exceeding one suggest rapid exponential growth, they are generally rare (with the notable exceptions of recently introduced invasive populations, \textcolor{blue}{and locally adapted ones}), suggesting that our model predictions may be highly relevant for SA polymorphism in many real-world populations. In addition, we find that much of the parameter space where a male-beneficial allele \textcolor{blue}{fixes} is demographically inviable for populations whose intrinsic growth rates are close to one prior to invasion. This demographic consequence of masculinization has rarely been considered \citep{hitchcock2020gene}. Moreover, these findings complement recent theoretical and empirical studies indicating that SA selection is likely to be both condition dependent, and stronger in locally-adapted populations near the center of a species' range, where population growth rates are expected to be higher \citep{BergerEtAl2014, Connallon2015}.

@@

\textcolor{blue}{One particularly important limitation of our main analysis is that we modeled density-independent growth. Of course, in natural populations density-dependent processes like resource competition or competition for breeding sites will often prevent populations from growing (or shrinking) exponentially. In the simple density-dependent version of our model presented in \ref{App:DensDep}, we included density-dependence acting on the survival of all life history stages equally, and reanalysed our case study of \textit{M. guttatus}. The density-dependent version of the model predicts equilibrium population densities rather than population intrinsic growth rates (see fig.~ \ref{fig:mimulusFig_dd}). This brief example shows that the shape of the polymorphic region is unchanged by this form of density-dependence, and that the shape of the population density surface in fig.~\ref{fig:mimulusFig_dd} is very similar to that for population growth rate in fig.~\ref{fig:mimulusFig}. The similarity between the density-independent and -dependent model predictions underscores an important point: density-dependence that acts (st)age-independently does not alter selection gradients \citep{mylius1995evolutionarily, caswell2017}.}

@@

\textcolor{blue}{Here we extend the model presented in the main text by incorporating density-dependent population regulation. We chose the simplest population regulation possible in a structured population model: population regulation affecting all life history stages in the same way. Note that this is the only type of population regulation that is possible in unstructured population models, like the classical Lotka-Volterra or Rosenzweig-MacArthur models. We model survival of all life-history stages as a exponentially decreasing function of population density, such that the population projection becomes
\begin{linenomath*}
\begin{equation}
	\tilde{\mbf{n}}(t + 1) = \exp\left(-\alpha \sum_i n_i \right) \tilde{\mbf{A}}[\tilde{\mbf{n}}(t)] \, \tilde{\mbf{n}}(t),
\end{equation}
\end{linenomath*}
where $\sum_i n_i$ sums over all entries of the population state vector, and hence over all genotypes in all life-history stages. Equivalently, in matrix algebra notation,
\begin{linenomath*}
\begin{equation}
	\tilde{\mbf{n}}(t + 1) = \exp\left(-\alpha \| \tilde{\mbf{n}}(t) \|\right) \tilde{\mbf{A}}[\tilde{\mbf{n}}(t)] \, \tilde{\mbf{n}}(t),
\end{equation}
\end{linenomath*}
where $\| \tilde{\mbf{n}}(t) \|$ is the one-norm which is identical to the sum of the entries of $\tilde{\mbf{n}}(t)$ because the population vector only has nonnegative entries. }

>>>>

%To persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1937}, \textcolor{blue}{rapid, unpredictable, or extreme changes} in environmental conditions \citep{Maynard-Smith1976, LandeShannon1996, OrrUnckless2008,Lyberger-etal-2021}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when \textcolor{blue}{different fitness components are negatively correlated} \citep{CharlesworthHughes2000, ConnallonHall2018}. 
\textcolor{blue}{A population's ability to adapt to its environment and persist in the long term depends on the genetic variation that it carries \citep{Fisher1930, Lewontin1975}. Yet, populations are heterogenous collections of individuals existing in a complex world where adaptation can be impeded by conflicting selection and gene-flow between different classes of individuals, or more generally, when different fitness components are negatively correlated \citep{Prout2000, CharlesworthHughes2000, ConnallonHall2018}.}

@@

\textcolor{blue}{While significant efforts have been made to predict when SA polymorphisms should be maintained and to detect SA genetic variation from empirical data, less attention has been paid to the demographic consequences of SA selection \citep{MatthewsConnallon2019}}. Most population genetic theory, including SA selection theory, models the relative fitness of genotypes in populations of constant (and often infinite) size. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals \textcolor{blue}{which link genotype distributions to population dynamics (e.g., \citealt {johnston2013life, merot2020balancing}). For over 50 years, the field of eco-evolutionary dynamics has sought to jointly model the genetic and demographic processes involved in evolutionary change. Theoretical models combining population genetics with density-dependent population growth \citep{roughgarden1971density} and stage-structured populations (consolidated in \citealt{Charlesworth_1994}) paved the way toward understanding the demographic consequences of allele frequency changes due to selection. Various extensions of these theories have been made to study life-history evolution \citep{orive1995senescence, orive2001somatic}, spatial structure \citep{ronce2000kin}, kin selection \citep{rousset2004genetic}, and the evolution of quantitative traits \citep[e.g.,][]{coulson2006putting, barfield2011evolution, childs2016evolution, orive2017effects}. This growing body of eco-evolutionary theory makes plain that demographic processes must be taken into account explicitly to fully understand how genetic variation for fitness is shaped, and maintained in natural populations \citep{metcalf2007evolutionary,coulson2006putting,rueffler2006evolutionary}.}

@@

\textcolor{blue}{Individuals contribute  gametes to the population outcrossing gamete pool according to their fertility, which depends on their stage and genotype at the SA locus. The allele frequencies in the male gamete pool are then calculated as 
\begin{linenomath*}
\begin{equation} \label{eq:maleGametePool}
	\left(
		\begin{array}{c}
			q^{\prime}_{A} \\
			q^{\prime}_{a} \\
		\end{array} \right) \propto 
				\left(
		\begin{array}{c}
			\mbf{1}^{\intercal}_{\omega}\mbf{F}^{\prime}_{AA}\mbf{p}^{\prime}_{AA}+ \frac{1}{2} \mbf{1}^{\intercal}_{\omega}  \mbf{F}^{\prime}_{Aa} \mbf{p}^{\prime}_{Aa}  \\
		\frac{1}{2} \mbf{1}^{\intercal}_{\omega}  \mbf{F}^{\prime}_{Aa} \mbf{p}^{\prime}_{Aa} + \mbf{1}^{\intercal}_{\omega}\mbf{F}^{\prime}_{aa}\mbf{p}^{\prime}_{aa}\\
		\end{array} \right),
\end{equation}
\end{linenomath*}
where equality is obtained by normalizing the frequencies such that they add to one (see Online Supplementary Materials for a derivation of Eq(\ref{eq:maleGametePool})). The first entry sums the  stage-specific contributions of allele A to the gamete pool by heterozygotes and $AA$ homozygotes; the second entry sums the stage-specific contributions of allele a to the gamete pool by heterozygotes and $aa$ homozygotes. }

\textcolor{blue}{ We now have all the ingredients to calculate the allele frequencies in the male gamete pool for the two-stage model in Eq(\ref{eq:F_male}), Eq(\ref{eq:maleGametePool}), and Table \ref{tab:Fitness}. Denoting the proportion of reproducing adults of genotype $i$ with $p^R_i$, the allele frequencies in the male gamete pool are given by 
\begin{linenomath*}
\begin{equation} \label{eq:maleGametePool_example}
\left(
		\begin{array}{c}
			q^{\prime}_{A} \\
			q^{\prime}_{a} \\
		\end{array} \right) = 
				\frac{1}{(1-s_m)p^R_{aa}+ (1-hs_m)p^R_{Aa}+p^R_{aa}}\left(
		\begin{array}{c}
(1-s_m)p^R_{AA}	+\frac{1}{2}(1-hs_m)p^R_{Aa}\\
p^R_{aa}+\frac{1}{2}(1-hs_m)p^R_{Aa}
		\end{array} \right).
\end{equation}
\end{linenomath*}
Note that this equation is identical to the classical population genetic result for allele frequencies after selection among males in a two-sex SA model, with the caveat that only reproductively active adults contribute to the gamete pool \citep[e.g., see Eq(1a) of][]{Kidwell1977}. Eq(\ref{eq:maleGametePool_example}) increasingly deviates from the classical population genetics result if additional reproductive adult age or stage classes are included (see Eq(\ref{eq:maleGametePool})).  }

@@

We identify conditions where SA polymorphism is 'protected' by evaluating the stability of populations initially fixed for either SA allele to invasion by the other allele (i.e., we assessed stability at the boundary equilibrium genotype frequencies of $\hat{\mbf{p}}_{AA} = 1$ and $\hat{\mbf{p}}_{aa} = 1$; \citealt{Levene1953, Prout1968, deVriesCaswell2019b}). The formal conditions for a protected polymorphism are determined by linearizing the model in the vicinity of the boundary equilibria ($\hat{\mbf{p}}$), and evaluating the magnitude of the largest eigenvalue of the Jacobian matrix of the linearization. A full derivation of the Jacobian and details of the invasion analysis are provided in the Online Supplementary Material, and the relevant leading eigenvalues are presented in \ref{App:Eigen}.

@@

Mortality from inbreeding depression quickly outweighs the beneficial effect of selfing on the equilibrium female load that was apparent in the previous section. In contrast to our earlier results, predominantly outcrossing populations are predicted to have the highest proportion of demographically viable polymorphic space when inbreeding depression is taken into acount (compare fig.~\ref{fig:polySpace} with fig.~\ref{fig:deltaPolySpace}). Regardless of the life-history stage at which inbreeding depression affects survival, populations with intermediate to high selfing rates are unlikely to harbour SA polymorphism unless they can afford the resulting loss of self-fertilized ovules/offspring. An interesting alternative interpetation of these results is that populations with intermediate to high selfing rates and population intrinsic growth rates near one are vulnerable to extinction if a sexually antagonistic allele invades the population.

@@

The first and central finding of our study is that when population intrinsic growth rates approach one, the deleterious effects of segregating male-beneficial SA alleles on female fecundity can result in extinction over much of the parameter space where SA polymorphism is maintained. We find that much of the parameter space where a male-beneficial allele \textcolor{blue}{fixes} is demographically inviable for populations whose intrinsic growth rates are close to one prior to invasion. Since intrinsic growth rates far exceeding one suggest rapid exponential growth, they are generally rare (with the notable exceptions of recently introduced invasive populations), suggesting that our model predictions may be highly relevant for SA polymorphism in many real-world populations. This demographic consequence of masculinization has rarely been considered \citep{hitchcock2020gene}. Moreover, these findings complement recent theoretical and empirical studies indicating that SA selection is likely to be both condition dependent, and stronger in locally-adapted populations near the center of a species' range, where population growth rates are expected to be higher and populations might be more likely to be able to suffer the demographic cost of a SA polymorphism while remaining viable \citep{BergerEtAl2014, Connallon2015}.

@@

\textcolor{blue}{One particularly important limitation of our main analysis is that we modeled density-independent growth. Of course, in natural populations density-dependent processes like resource competition or competition for breeding sites will often prevent populations from growing (or shrinking) exponentially. In the simple density-dependent version of our model presented in \ref{App:DensDep}, we included density-dependence acting on the survival of all life history stages equally, and reanalysed our case study of \textit{M. guttatus}. The density-dependent version of the model predicts equilibrium population densities rather than population intrinsic growth rates (see fig.~\ref{fig:mimulusFig_dd}). This brief example shows that the shape of the polymorphic region is unchanged by this form of density-dependence, and that the shape of the population density surface in fig.~\ref{fig:mimulusFig_dd} is very similar to that for population growth rate in fig.~\ref{fig:mimulusFig}. The similarity between the density-independent and -dependent model predictions underscores an important point: density-dependence that acts (st)age-independently does not alter selection gradients \citep{mylius1995evolutionarily, caswell2017senescence}.}

@@

\textcolor{blue}{Here we extend the model presented in the main text by incorporating density-dependent population regulation. We chose the simplest population regulation possible in a structured population model: population regulation affecting all life history stages in the same way. Note that this is the only type of population regulation that is possible in unstructured population models, like the classical Lotka-Volterra or Rosenzweig-MacArthur models. We model survival of all life-history stages as a exponentially decreasing function of population density, such that the population projection becomes
\begin{linenomath*}
\begin{equation}
	\tilde{\mbf{n}}(t + 1) = \exp\left(-\alpha \sum_i n_i \right) \tilde{\mbf{A}}[\tilde{\mbf{p}}(t)] \, \tilde{\mbf{n}}(t),
\end{equation}
\end{linenomath*}
where $\sum_i n_i$ sums over all entries of the population state vector, and hence over all genotypes in all life-history stages. Equivalently, in matrix algebra notation,
\begin{linenomath*}
\begin{equation}
	\tilde{\mbf{n}}(t + 1) = \exp\left(-\alpha \| \tilde{\mbf{n}}(t) \|\right) \tilde{\mbf{A}}[\tilde{\mbf{p}}(t)] \, \tilde{\mbf{n}}(t),
\end{equation}
\end{linenomath*}
where $\| \tilde{\mbf{n}}(t) \|$ is the one-norm which is identical to the sum of the entries of $\tilde{\mbf{n}}(t)$ because the population vector only has nonnegative entries. }
